---
title: "Email Campaign Newsletter Analysis"
author: "Nik8x"
date: "July 12, 2018"
output: rmarkdown::github_document
---
Email Campaign Newsletter

```{r}
subdata <- read.csv(file.choose(), header = T, na.strings = c("NA",""," ",".")) # choose the subscribers data csv file
```

```{r}
nrow(subdata)
ncol(subdata)

head(subdata)
```

```{r}
summary(subdata)
```

```{r}
str(subdata)
```

```{r}
length(unique(subdata$Profile.Id))
```
We can consider each Profile.Id to be a unique Subscriber, hence in total we have 705144 subscribers

Convert Signup(Date, Time) to Date, changing data type from factor to date 
```{r}
subdata$Date <- as.Date(subdata$Signup) # storing in a new column 'Date'
str(subdata$Date)
```

***
***

```{r}
# Let's break the Date column into Year
subdata$Year <- as.numeric(format(subdata$Date, "%Y" ))
str(subdata$Year)

# install.packages('plyr')
library(plyr)
yearcount <- count(subdata, 'Year')
yearcount
```

```{r}
# Plotting above data

# install.packages('ggplot2')
library(ggplot2)

ggplot(yearcount, aes(x = Year, y = freq)) + 
  geom_bar(aes(fill = freq), stat = "identity") +
  xlab("Year") + ylab("Number of Subscribers") + labs(title = "Number of Subscribers per Year") +
  geom_text(aes(label = freq), vjust = -1)
```
We got most subscribers in the year 2016

```{r}
# Let's see referral sources
count(subdata, 'referral_source')
```

We can see that bookbub, Bookbub; Bookpage, BookPage; bookriot, Bookriot; google, Google; facebook, Facebook, facebook.com; litbreaker, Litbreaker; 
Pinterest, pinterest.com; The Line Up, The Line Up, The Lineup ; and twitter, Twitter are counted  separately per year, let's rename them
```{r}
subdata$referral_source[subdata$referral_source == 'bookbub'] <- 'Bookbub'
subdata$referral_source[subdata$referral_source == 'Bookpage'] <- 'BookPage'
subdata$referral_source[subdata$referral_source == 'bookriot'] <- 'Bookriot'
subdata$referral_source[subdata$referral_source == 'google'] <- 'Google'
subdata$referral_source[subdata$referral_source == 'facebook'] <- 'Facebook'
subdata$referral_source[subdata$referral_source == 'facebook.com'] <- 'Facebook'
subdata$referral_source[subdata$referral_source == 'litbreaker'] <- 'Litbreaker'
subdata$referral_source[subdata$referral_source == 'pinterest.com'] <- 'Pinterest'
subdata$referral_source[subdata$referral_source == 'The Line Up '] <- 'The Lineup'
subdata$referral_source[subdata$referral_source == 'The Line Up'] <- 'The Lineup'
subdata$referral_source[subdata$referral_source == 'twitter'] <- 'Twitter'
```


Count	number	of	subscribers	we	acquired	each	year	by	referral source
```{r}
y <- aggregate(Profile.Id ~ referral_source + Year, data = subdata, length)
```


```{r}
y1 <- y[with(y,order(-Profile.Id)),]
y1 <- y1[1:10,]
y1
```

```{r}
ggplot(y1, aes(x = referral_source, y = Profile.Id)) + 
  geom_bar(aes(fill = y1$Year), stat = "identity") +
  xlab("Source") + ylab("Number of Subscribers") + labs(title = "Number of Subscribers per Year") +
  #geom_text(aes(label = Profile.Id), vjust = -1) +
  theme(axis.text.x = element_text(angle = 65, vjust = 0.6))
```


```{r}
# Let's break the Date column into Month
subdata$Month <- as.factor(format(subdata$Date, "%B" ))
```

Count	number	of	subscribers	we	acquired	each	month	by	referral source
```{r}
z <- aggregate(Profile.Id ~ referral_source + factor(Month, levels = month.name), data = subdata, length)
```

```{r}
z1 <- z[with(z,order(-Profile.Id)),]
z1 <- z1[1:10,]
z1
```


```{r}
ggplot(z1, aes(x = referral_source, y = Profile.Id)) + 
  geom_bar(aes(fill = z1$`factor(Month, levels = month.name)`), stat = "identity") +
  xlab("Source") + ylab("Number of Subscribers") + labs(title = "Number of Subscribers per Month") +
  #geom_text(aes(label = Profile.Id), vjust = -1) +
  theme(axis.text.x = element_text(angle = 65, vjust = 0.6))
```
In the month of January, DMi gave us 43393 suscribers

```{r}
x <- aggregate(Profile.Id ~ referral_source + Year + factor(Month, levels = month.name), data = subdata, length)
```

```{r}
x1 <- x[with(x,order(-Profile.Id)),]
x1 <- x1[1:10,]
x1
```
43233 subscribers were acquired in the month of January, 2016, via DMi.

***
***

Let us create a new data frame consisting of subscriber and their preferences
```{r}
attach(subdata)
subdata_ebb <- data.frame(Profile.Id, ebb_preferences)
detach(subdata)
```

```{r}
# Now let's remove square brackets and quotations from preferences

subdata_ebb$ebb_preferences <- paste0("", gsub("[][]", "", subdata_ebb$ebb_preferences), "") # removing square brackets

subdata_ebb$ebb_preferences <- paste0("", gsub('""', "", subdata_ebb$ebb_preferences), "") # removing quotations
```

```{r}
subdata_ebb$ebb_preferences <- sapply(subdata_ebb$ebb_preferences, function(x) gsub("\"", "", x)) # remove \" from preferences 

subdata_ebb$ebb_preferences <- as.character(subdata_ebb$ebb_preferences) # store this as character
```

```{r}
# Let's see total count of preferences per subscriber
library(stringi)
subdata_ebb$ebb_preferences_count <- stri_count_words(subdata_ebb$ebb_preferences)
```

```{r}
# The missing values should be replaced with maximum preferences number, let's find that
summary(as.factor(subdata_ebb$ebb_preferences_count))
```
We see that most of the preferences are 14(only 2 values for 16 so we exclude it)

```{r}
# Fill empty values in preferences with 14
subdata_ebb$ebb_preferences_count[subdata_ebb$ebb_preferences == 'NA'] <- 14
```

```{r}
# plot above data

ggplot(subdata_ebb, aes(ebb_preferences_count)) +
  geom_histogram() +
  xlab("Preferences") + ylab("Number of Subscribers") + labs(title = "Number of Subscribers")
```
Most subscribers are subscribing to all 14 preferences. It might be because most people prefer default selection.

***
***

```{r}
# Let's store preferences into a list
preferences <- strsplit(subdata_ebb$ebb_preferences, split = ",", fixed = TRUE)
# head(preferences)

preferences_unlist <- unlist(preferences)
# head(preferences_unlist, 10)

unique(preferences_unlist)
```


We have repetitions here, Let us write down the above list without repetitions and significant preferences
bestsellers
mysteries_thrillers_true_crime
science_fiction_fantasy
literary_fiction
general_nonfiction
book_club_picks
historical_fiction
romance 
young_adult
biographies_and_memoir
lifestyle_and_cooking
childrens
history
poetry


We can also see that there are characters y, g, s, p, r, c which in themself do not signify any preference, so we check and drop them
```{r}
which(preferences_unlist == "y")
which(preferences_unlist == "g")
which(preferences_unlist == "s")
which(preferences_unlist == "p")
which(preferences_unlist == "r")
which(preferences_unlist == "c")
```
There are not a whole lot of them, so we will notinclude them

Let us match the pattern in preferences and encode them into binary columns
```{r}
subdata_ebb$bestsellers <- ifelse(grepl("bestsellers", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$mysteries <- ifelse(grepl("mysteries", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$science_fiction <- ifelse(grepl("sc", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$lit_fiction <- ifelse(grepl("literary_fic", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$general_nonfiction <- ifelse(grepl("ge", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$book_club_picks <- ifelse(grepl("book_club_picks", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$historical_fiction <- ifelse(grepl("hist", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$romance <- ifelse(grepl("ro", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$young_adult <- ifelse(grepl("yo", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$bio_memoirs <- ifelse(grepl("bio", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$lifestyle_cooking<- ifelse(grepl("life", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$childrens <- ifelse(grepl("ch", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$history<- ifelse(grepl("history", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
subdata_ebb$poetry <- ifelse(grepl("po", subdata_ebb$ebb_preferences, fixed = TRUE),1,0)
```

```{r}
head(subdata_ebb)
```

***

### Predictive Model

Let us prepare our data
Let's again look at all the columns we got
```{r}
colnames(subdata)
colnames(subdata_ebb)
```
Domain and Engagement are irrelevant; Geolocation, Age, Gender, TopDevice, referral source have too many missing values, so we will not consider them

For predicting when people willl opt-out of the newsletter we will only be needing significant and related variables. Let's create a new dataframe and store all the significant columns which we are going to consider for prediction
```{r}
attach(subdata)
optoutprediction <- data.frame(subdata_ebb, Year, Date, Opens, Clicks, Lifetime.Message, Signup, Last.Open, Optout.Time)
detach(subdata)
```

Let's look at the amount of data missing in Optout Time
```{r}
sum(is.na(optoutprediction$Optout.Time))
```
We will drop these rows

```{r}
optoutprediction <- optoutprediction[!is.na(optoutprediction$Optout.Time), ]
```

```{r}
nrow(optoutprediction)
```
We are left with 186088 rows only

```{r}
optoutprediction$optoutduration <- difftime(optoutprediction$Optout.Time ,optoutprediction$Signup , units = c("days"))
optoutprediction$activeduration <- difftime(optoutprediction$Last.Open ,optoutprediction$Signup , units = c("days"))
```

```{r}
max(optoutprediction$optoutduration)
```

Also convert date into month
```{r}
optoutprediction$Month <- as.factor(format(optoutprediction$Date, "%m" ))
```

ebb_preferences, Date, Signup, Last.Open, Optout.Time is irrelevant, we will drop it
```{r}
optoutprediction$ebb_preferences <- NULL
optoutprediction$Date <- NULL
optoutprediction$Signup <- NULL
optoutprediction$Last.Open <- NULL
optoutprediction$Optout.Time <- NULL
```

Let's look at the structure of our new dataset
```{r}
str(optoutprediction)
```
Looks like we have to convert some of the data types

```{r}
optoutprediction$bestsellers <- as.factor(optoutprediction$bestsellers)
optoutprediction$mysteries <- as.factor(optoutprediction$mysteries)
optoutprediction$science_fiction <- as.factor(optoutprediction$science_fiction)
optoutprediction$lit_fiction <- as.factor(optoutprediction$lit_fiction)
optoutprediction$general_nonfiction <- as.factor(optoutprediction$general_nonfiction)
optoutprediction$book_club_picks <- as.factor(optoutprediction$book_club_picks)
optoutprediction$historical_fiction <- as.factor(optoutprediction$historical_fiction)
optoutprediction$romance <- as.factor(optoutprediction$romance)
optoutprediction$young_adult <- as.factor(optoutprediction$young_adult)
optoutprediction$bio_memoirs <- as.factor(optoutprediction$bio_memoirs)
optoutprediction$lifestyle_cooking <- as.factor(optoutprediction$lifestyle_cooking)
optoutprediction$childrens <- as.factor(optoutprediction$childrens)
optoutprediction$history<- as.factor(optoutprediction$history)
optoutprediction$poetry <- as.factor(optoutprediction$poetry)
optoutprediction$Year <- as.factor(optoutprediction$Year)
optoutprediction$optoutduration <- as.numeric(optoutprediction$optoutduration)
optoutprediction$activeduration <- as.numeric(optoutprediction$activeduration)
```

Let's look again
```{r}
str(optoutprediction)
```

Before applying any model, let's check for missing values
```{r}
summary(optoutprediction)
```
optoutduration and activeduration have negative values which we will make positive, and activeduration has missing values

```{r}
optoutprediction$optoutduration <- abs(optoutprediction$optoutduration)
optoutprediction$activeduration <- abs(optoutprediction$activeduration)
```

```{r}
summary(optoutprediction$activeduration)
```
These missing values can be replace by median

```{r}
optoutprediction$activeduration[is.na(optoutprediction$activeduration)] <- median(optoutprediction$activeduration, na.rm = TRUE)
```

```{r}
summary(optoutprediction$activeduration) # Now it' good
```

#### Prediction

```{r}
label <- optoutprediction$optoutduration # our variable for prediction
```

```{r}
set.seed(1234)
oneortwo <- sample(1:2 , length(optoutprediction$Profile.Id), replace = TRUE, prob=c(0.8, 0.2)) # generating random values and storing them
```

```{r}
# create train data frame
train <- optoutprediction[oneortwo == 1, ]

# create test data frame
test <- optoutprediction[oneortwo == 2, ]
```

##### Let's apply Linear Regression model
```{r}
lm_fit <- step(lm(optoutduration ~. -Profile.Id, data = train), direction = "both")
```

```{r}
summary(lm_fit)
```
Adjusted R-squared:  0.9471 which is really good, and nearly all our variables are significant


We will now apply the prediction model to the test data.
```{r}
predictedvalues <- predict(lm_fit, newdata = test)
```

Now, let’s look at the first few values of prediction, and compare it to the values of optoutduration in the test data set
```{r}
head(predictedvalues)
```

```{r}
head(test$optoutduration)
```

This means that for the value of 500.028148, our prediction was 462.27999, for 568.744468, it was 639.89648, for 44.306655, it was 52.19734 and so on.

Baseline model
```{r}
bestguess <- median(train$optoutduration) 
```

```{r}
RMSE.baseline <- sqrt(median((bestguess - train$optoutduration) ^ 2))
RMSE.baseline
```


Evaluate the accuracy
```{r}
RMSE.predictedvalues <- sqrt(mean((predictedvalues - test$optoutduration) ^ 2))
RMSE.predictedvalues
```

Apparently our's is good model but can be a lot better

##### Decision Tree
```{r}
# Needed to grow a tree
library(rpart)
#To draw a pretty tree 
library(rattle)
```


```{r}
rt <- rpart(optoutduration ~ . , data = train[-1], method = "anova")
```

```{r}
# Examine life_predicted_2
plot(rt)
text(rt)
```

```{r}
fancyRpartPlot(rt) # This gives better plot
```

```{r}
predicted_dt <- predict(rt,test)
```

Now, let’s look at the first few values of prediction, and compare it to the values of optoutduration in the test data set
```{r}
head(predicted_dt)
```

```{r}
head(test$optoutduration)
```

Evaluate the accuracy
```{r}
RMSE.predicted_dt <- sqrt(mean((predicted_dt - test$optoutduration) ^ 2))
RMSE.predicted_dt
```

##### I think it would be better to build a classification model by classifying optoutduration variable in Quarter monthly duration.